---
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


#ecostattoolkit

An R package to implement and test the ECOSTAT toolkit for estimating nutrient boundaries.

Implements and tests the ECOSTAT toolkit for estimating nutrient
boundaries. R functions have been created that replicate the original scripts by
Geoff Phillips.


## Installation

ecostattoolkit can be installed directly from GitHub using the `install_github` function in package `devtools`

```{r install, eval = FALSE}
# install.packages("devtools")
devtools::install_github("andrewdolman/ecostattoolkit")
```


## Usage

### Required packages

```{r, message=FALSE, warning=FALSE}
library(ecostattoolkit)
library(dplyr)
library(tidyr)
```

### Read in and prepare data
```{r}
filename <- system.file("extdata", "DataTemplate_Example1.csv", package = "ecostattoolkit")
dat <- read.csv(filename)

status_boundaries <- c("HG" = 0.8, "GM" = 0.6, "MP" = 0.4)

# min and max nutrient value used for model
nut.min <- 1	
nut.max <- 138	

dat$status <- cut(dat$EQR,
                  breaks = rev(c(Inf, status_boundaries, -Inf)),
                  labels = rev(c("High", "Good", "Moderate", "Poor")))

dat$x.u <- log10(dat$P)
dat$y.u <- dat$EQR

knitr::kable(head(dat))
```

### Estimate boundaries with all methods
```{r}
dat_sub <- filter(dat, Exclude_P == 0,
                  P <= nut.max,
                  P >= nut.min)

boundaries <- boundaries_all(dat)


boundaries <- dplyr::mutate_if(boundaries, is.numeric, dplyr::funs(round(10^., 0))) %>% 
  select(-MP) 

knitr::kable(boundaries)

```

### Compare with results from scripts in *TKit_20170208_CORRECTED*

Run script

```{r, fig.width=5}
source(system.file("extdata",
                   "TKit_20170208_CORRECTED/TKit_fit_lin_mod1.R",
                   package = "ecostattoolkit"))
```


```{r gather_script_results}
# Collect results
script.res <- data.frame(
  Method = c(rep("OLS1_YonX", 2), rep("OLS2_XonY", 2),
             rep("RMA", 2)),
  
  Boundary = c("GM", "HG", "GM", "HG",
               "GM", "HG"),
  
  Value = c(GM.mod1, HG.mod1, GM.mod2, HG.mod2,
            GM.mod4, HG.mod4),
  stringsAsFactors = TRUE)

script.res %>% 
  tidyr::spread(Boundary, Value) %>% 
  dplyr::select(Method, HG, GM) %>% 
  arrange(Method) %>% 
  knitr::kable(.)

```