---
title: "Comparison of methods by simulation"
author: "Andrew M. Dolman"
date: "27 September 2016"
output: pdf_document
---

```{r setup, echo = FALSE, message=FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)
#library(ordinal)
library(purrr)
library(mgcv)
library(ecostattoolkit)

knitr::opts_chunk$set(tidy=TRUE, warning=FALSE, echo=FALSE, dpi = 300, cache = T, autodep = T, fig.height = 8)
```

```{r create_simulated_data}
EQR_boundaries <- data.frame(Boundary = c("HG", "GM", "MP"), Value = c(0.80, 0.60, 0.40))

log10_TP_slope <- -1
log10_TP_intercept <- -(log10(50) * log10_TP_slope - EQR_boundaries[EQR_boundaries$Boundary=="GM", "Value"])

set.seed(2)

get_sim_dat <- function(n, xerr = FALSE, slope = -1, boundaries = EQR_boundaries){
  
  log10_TP_slope <- slope
  log10_TP_intercept <- -(log10(50) * log10_TP_slope - boundaries[boundaries$Boundary=="GM", "Value"])

  sim_dat <- data.frame(Region = rep(c("Few low nutrient lakes", "Good data span", "Few high nutrient lakes"),
                                     each = n),
                        TP = c(
                          10 ^ rnorm(n, log10(100), 0.25),
                          10 ^ rnorm(n, log10(50), 0.25),
                          10 ^ rnorm(n, log10(35), 0.25))
  ) %>%
    mutate(log10_TP = log10(TP),
           EQR = log10_TP_intercept + log10_TP_slope * log10(TP) + rnorm(n, 0, 0.3)) %>%
    mutate(Bio_Class = cut(EQR, breaks = c(-Inf, boundaries$Value, Inf),
                           labels = rev(c("High", "Good", "Moderate", "Poor")),
                           ordered_result = TRUE),
           Bio_Class_2 = cut(EQR, 
                             c(-Inf, boundaries[boundaries$Boundary=="GM","Value"], Inf),
                             labels = c("Mod_or_Worse", "Good_or_better")),
           Bio_Class_num = as.numeric(Bio_Class)+1) %>%   tbl_df() %>% 
    mutate(log10_TP  = if(xerr){log10_TP + rnorm(n, 0, xerr)}else{log10_TP},
           TP = 10^log10_TP)
  return(sim_dat)
}

sim_dat <- get_sim_dat(n = 100, xerr = F)
```



```{r}
p <- sim_dat %>% 
  ggplot(aes(x = TP)) %>% 
  + geom_histogram(bins = 15) %>% 
  + facet_wrap( ~ Region, ncol = 1) %>% 
  + scale_x_continuous(trans = "log10", breaks = c(1, 5, 10, 50, 100, 500, 1000)) %>% 
  + annotation_logticks(sides = "b")
p


```

```{r}
p <- sim_dat %>% 
  ggplot(aes(x = EQR)) %>% 
  + geom_histogram(bins = 15) %>% 
  + facet_wrap( ~ Region, ncol = 1) #%>% 
p
```

```{r}
p <- sim_dat %>% 
  ggplot(aes(x = TP, y = EQR)) %>% 
  + geom_point() %>% 
  + geom_smooth(method = "lm") %>% 
  + facet_wrap( ~ Region, ncol = 1) %>% 
  + scale_x_continuous(trans = "log10", breaks = c(1, 5, 10, 50, 100, 500, 1000)) %>% 
  + annotation_logticks(sides = "b") %>% 
  + geom_hline(data = filter(EQR_boundaries, Boundary == "GM"), aes(yintercept = Value, colour = Boundary)) %>% 
  + geom_vline(data = filter(EQR_boundaries, Boundary == "GM"), aes(xintercept = 50, colour = Boundary))
p
```


```{r estimate_TP_boundaries}
get_TP_boundary_lm <- function(lm_mod){
  cfs <- coef(lm_mod)
  10^((EQR_boundaries[EQR_boundaries$Boundary=="GM","Value"] - cfs[1]) / cfs[2])
}

lm_test <- lm(EQR ~ log10_TP, data = sim_dat)
#get_TP_boundary_lm(lm_test)

get_TP_boundary_gam_ord <- function(gam_ord_mod){
  newdat = expand.grid(log10_TP = seq(log10(5), log10(200), length.out = 200))
  newdat <- bind_cols(newdat, data.frame(predict(gam_ord_mod, type="response", se = F, newdata = newdat)))
  newdat <- newdat %>% 
    rename(Poor = X1, Mod = X2, Good = X3, High = X4) %>% 
    mutate(diff_HG_MP = (High + Good - Mod - Poor)^2,
           diff_GM = (Good - Mod)^2) %>% 
    summarise(`GAM_ORD_GM` = 10^(log10_TP[which.min(diff_HG_MP)]))#,
              #`gam_ord_G-M` = 10^(log10_TP[which.min(diff_GM)]))
  return(newdat)
}

gam_test <- gam(I(Bio_Class_num-1) ~ log10_TP, family=ocat(R=4), data = sim_dat)

#get_TP_boundary_gam_ord(gam_test)
```

```{r mismatch_method}
get_Mismatch_GM_2 <- function(data){
  newdat <- expand.grid(TP = seq(min(data$TP), max(data$TP), length.out = 100)) %>%
    group_by(TP) %>%
    mutate(TP.not.Good_EQR_Good = sum(data$TP > TP & data$EQR >= 0.6) ,
           TP.Good_EQR.not.Good = sum(data$TP < TP & data$EQR < 0.6) ,
           Mismatch_diff = abs(TP.not.Good_EQR_Good - TP.Good_EQR.not.Good)) %>%
    ungroup() 
  plot(TP.not.Good_EQR_Good~TP, data = newdat, type = "l")
  lines(TP.Good_EQR.not.Good~TP, data = newdat, col =  "Red")
  rug(data$TP)
   
  newdat <- newdat %>%
    summarise(Mismatch_GM = TP[which.min(Mismatch_diff)])
    
  return(newdat)
}

get_Mismatch_GM <- function(data){
  f <- function(x, data){
    TP.not.Good_EQR_Good = sum(data$TP > exp(x) & data$EQR >= 0.6, na.rm = T)
    TP.Good_EQR.not.Good = sum(data$TP < exp(x) & data$EQR < 0.6, na.rm = T)
    #ratio = (sum(data$TP < 50)) / sum((data$TP >= 50)) 
    Mismatch_diff = (TP.not.Good_EQR_Good - TP.Good_EQR.not.Good)^2
    return(Mismatch_diff)
  }
  exp(optimize(f, c(log(min(data$TP)), log(max(data$TP))), data = data)$minimum)
}

#
# sim_dat <- get_sim_dat(n = 5, xerr = 0)
# get_Mismatch_GM(sim_dat)
# get_Mismatch_GM_2(sim_dat)
# 
# sim_dat %>% 
#   group_by(Region) %>% 
#   summarise(B = get_Mismatch_GM_2(.)$Mismatch_GM)

```


```{r}
get_quantile_methods <- function(data){
  data %>% 
    group_by(Bio_Class) %>% 
    summarise(q25 = quantile(TP, c(0.25)),
              q50 = quantile(TP, c(0.5)),
              q75 = quantile(TP, c(0.75))) %>% 
    gather(quant, Value, -Bio_Class) %>% 
    unite(Bio_Class_quant, Bio_Class, quant, sep = "_") %>% 
    spread(Bio_Class_quant, Value) %>% 
    summarise(Adj_q_GM = mean(c(Good_q75, Moderate_q25), na.rm = T),
              q50_G = Good_q50,
              q75_G = Good_q75)
}

#
sim_dat <- get_sim_dat(n = 30, xerr = 0.3)
#get_quantile_methods(sim_dat)

```


```{r sim_err_x}
get_estimates <- function(data){
  
  method_types = c(OLS_GM = "Regression", MM_GM = "Mismatch", q50_G = "Quantile",
                   q75_G = "Quantile", Adj_q_GM = "Quantile")
  
  ests <- data %>% 
    nest(-Region, -Rep) %>% 
    mutate(lm_fit = map(data, ~ lm(EQR ~ log10_TP, data = .)),
           #glm_fit = map(data, ~ glm(Bio_Class_2 ~ log10_TP, data = ., family = binomial)),
           #gam_ord_fit = map(data, ~ gam(I(Bio_Class_num-1) ~ log10_TP, family=ocat(R=4), data = .)),
           OLS_GM = map(lm_fit, get_TP_boundary_lm),
           #GLM_GM = map(glm_fit, ~10^MASS::dose.p(., p = 0.5)[1]),
           #GAM_ORD_GM = map(gam_ord_fit, get_TP_boundary_gam_ord),
           MM_GM = map(data, ~get_Mismatch_GM(.)),
           quantile_methods = map(data, ~get_quantile_methods(.)))%>% 
    unnest(OLS_GM, MM_GM, quantile_methods) %>% 
    select(-data, -lm_fit) %>% 
    gather(Method, Value, -Region, -Rep) %>% 
    mutate(Method_type = method_types[match(Method, names(method_types))],
           Method = factor(Method, levels = c("q50_G", "q75_G", "Adj_q_GM", "OLS_GM", "MM_GM"), ordered = T))
  
  return(ests)
}

```

```{r}
plot_sim_out <- function(dat){
  p <- dat %>% 
    mutate(Value = ifelse(Value > 1000, Inf, Value)) %>% 
  ggplot(aes(x = Method, y = Value, fill = Method_type)) %>% 
  + geom_hline(yintercept = 50) %>% 
  + geom_boxplot() %>% 
  + facet_wrap(~ Region, nrow = 2) %>% 
  + labs(y = "Target nutrient concentration") %>% 
    + expand_limits(y = 0)
return(p)
}
```


## With measurement error only in the EQR

```{r}
dat <- get_sim_dat(60, xerr = F)

status_boundaries <- c("HG" = 0.8, "GM" = 0.6, "MP" = 0.4)

boundaries <- boundaries_all(dat, xvar = "log10_TP", yvar = "EQR",
                             status_boundaries = status_boundaries,
                             class_var = "Bio_Class")

boundaries <- dat %>% 
  group_by(Region) %>% 
  do(boundaries_all(., xvar = "log10_TP", yvar = "EQR",
                             status_boundaries = status_boundaries,
                             class_var = "Bio_Class"))

boundaries %>% 
  select(Region, Method, GM) %>% 
  spread(Method, GM) %>% 
  mutate_if(is.numeric, funs(round(10^.)))
```


```{r}
set.seed(1)
sim_out_errY <- plyr::ldply(1:100, function(x) get_sim_dat(60, xerr = F) %>%
                              mutate(Rep = x)) %>% 
  tbl_df() %>%
  get_estimates(.) 

plot_sim_out(sim_out_errY)
```


## With 1/2 (proportional) observation error in nutrient measurements compared to EQR 


```{r}
sim_out_errX0.5Y <- plyr::ldply(1:100, function(x) get_sim_dat(60, xerr = 0.15) %>%
                              mutate(Rep = x)) %>%
  tbl_df() %>%
  get_estimates(.) 

plot_sim_out(sim_out_errX0.5Y)
```




## With equal (proportional) observation error in nutrient and EQR measurements


```{r}
sim_out_errXY <- plyr::ldply(1:100, function(x) get_sim_dat(60, xerr = 0.3) %>%
                              mutate(Rep = x)) %>%
  tbl_df() %>%
  get_estimates(.) 

plot_sim_out(sim_out_errXY)
```


## With extra (proportional) observation error in nutrient and EQR measurements


```{r}
sim_out_errXXY <- plyr::ldply(1:100, function(x) get_sim_dat(60, xerr = 0.5) %>%
                              mutate(Rep = x)) %>%
  tbl_df() %>%
  get_estimates(.) 

plot_sim_out(sim_out_errXXY)
```



# Non-linear relationships (after logging)


```{r}
sim_out_nonlinear <- plyr::ldply(1:100, function(x) get_sim_dat(60, xerr = 0) %>%
                              mutate(Rep = x, 
                                     EQR = ifelse(EQR < 0, 0, EQR))) %>%
  tbl_df() %>%
  get_estimates(.) 

plot_sim_out(sim_out_nonlinear)
```


## Even less linear
```{r}
sim_out_nonlinear <- plyr::ldply(1:100, function(x) get_sim_dat(60, xerr = 0) %>%
                              mutate(Rep = x, 
                                     EQR = abs(EQR))) %>%
  tbl_df() %>%
  get_estimates(.) 

plot_sim_out(sim_out_nonlinear)
```


# Lower signal to noise
```{r}
sim_out_slope0 <- plyr::ldply(1:100, function(x) get_sim_dat(60, xerr = 0, slope = -0.5) %>%
                              mutate(Rep = x)) %>%
  tbl_df() %>%
  get_estimates(.) 

plot_sim_out(sim_out_slope0)
```


# No causal relationship
```{r}
sim_out_slope0 <- plyr::ldply(1:100, function(x) get_sim_dat(60, xerr = 0, slope = 0) %>%
                              mutate(Rep = x)) %>%
  tbl_df() %>%
  get_estimates(.) 

plot_sim_out(sim_out_slope0)
```


# Wild outliers


```{r}
sim_out_outliers1 <- plyr::ldply(1:100, function(x) get_sim_dat(60, xerr = 0, slope = -1) %>%
                              mutate(Rep = x, 
                                     EQR = EQR + rbinom(length(EQR), 1, 0) * rt(length(EQR), 2),
                                     Bio_Class = cut(EQR, breaks = c(-Inf, EQR_boundaries$Value, Inf),
                         labels = rev(c("High", "Good", "Moderate", "Poor")),
                         ordered_result = TRUE))) %>%
  tbl_df() %>%
  get_estimates(.) %>% 
  mutate(Outliers = "0%")


sim_out_outliers2 <- plyr::ldply(1:100, function(x) get_sim_dat(60, xerr = 0, slope = -1) %>%
                              mutate(Rep = x, 
                                     EQR = EQR + rbinom(length(EQR), 1, 0.2) * rt(length(EQR), 2),
                                     Bio_Class = cut(EQR, breaks = c(-Inf, EQR_boundaries$Value, Inf),
                         labels = rev(c("High", "Good", "Moderate", "Poor")),
                         ordered_result = TRUE))) %>%
  tbl_df() %>%
  get_estimates(.) %>% 
  mutate(Outliers = "20%")

sim_out <- bind_rows(sim_out_outliers1, sim_out_outliers2)

plot_sim_out(sim_out) %>%  
  + facet_wrap(~Region + Outliers, labeller = "label_both", ncol = 2)
```


## Wild outliers + slope -0-5

```{r}
sim_out_outliers1 <- plyr::ldply(1:100, function(x) get_sim_dat(60, xerr = 0, slope = -0.5) %>%
                              mutate(Rep = x, 
                                     EQR = EQR + rbinom(length(EQR), 1, 0) * rt(length(EQR), 2),
                                     Bio_Class = cut(EQR, breaks = c(-Inf, EQR_boundaries$Value, Inf),
                         labels = rev(c("High", "Good", "Moderate", "Poor")),
                         ordered_result = TRUE))) %>%
  tbl_df() %>%
  get_estimates(.) %>% 
  mutate(Outliers = "0%")


sim_out_outliers2 <- plyr::ldply(1:100, function(x) get_sim_dat(60, xerr = 0, slope = -0.5) %>%
                              mutate(Rep = x, 
                                     EQR = EQR + rbinom(length(EQR), 1, 0.2) * rt(length(EQR), 2),
                                     Bio_Class = cut(EQR, breaks = c(-Inf, EQR_boundaries$Value, Inf),
                         labels = rev(c("High", "Good", "Moderate", "Poor")),
                         ordered_result = TRUE))) %>%
  tbl_df() %>%
  get_estimates(.) %>% 
  mutate(Outliers = "20%")

sim_out <- bind_rows(sim_out_outliers1, sim_out_outliers2)

plot_sim_out(sim_out) %>%  
  + facet_wrap(~Region + Outliers, labeller = "label_both", ncol = 2)
```


